---
title: "Tidying Data"
author: "Alex Vand"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(palmerpenguins)
```

Read in the wb_indicators.csv file, make sure to specify `na = c("..", "")`

```{r}
wb_indicators <- read_csv(here::here("data", "wb_indicators.csv"), na = c("..", ""))
```

## Reshape so that year is a single column

```{r}
wb_indicators_long <- wb_indicators %>% 
  pivot_longer(cols = '2001 [YR2001]':'2020 [YR2020]',
               names_to = "year",
               values_to = "indicator_value")
```

## Clean up year column

```{r}
wb_data_sep <- wb_indicators_long %>% 
  separate(col = year, into = c("year", "year_chr"), sep = " ") %>% 
  select(-year_chr, -"Country Code", -"Series Code")
```

```{r}
df <- tribble(
  ~name, ~A, ~B, ~C,
  "teddy", 4, 5, 6,
  "khora", 9, 10, 11
)

df_long <- df %>% 
  pivot_longer(cols = A:C,
               names_to = "dog_food",
               values_to = "rankings")

df_wide <- df_long %>% 
  pivot_wider(names_from = dog_food, values_from = rankings)
```

```{r}
unique(wb_indicators_long$"Series Name")

wb_data_tidy <- wb_data_sep %>% 
  drop_na('Series Name') %>% 
  pivot_wider(names_from = 'Series Name', values_from = indicator_value)
          

names(wb_data_tidy) <- c("country", "year", "access_clean_fuels", "access_electricity", "co2_emissions", "fossil_fuel_cons", "water_stress")

wb_data_tidy <- wb_data_tidy %>% 
  rename(country_name = country,
         electricity = access_electricity)
```

## Practice some tidyverse functions for data wrangling

- `filter()` ROWS
- `select()` COLUMNS
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- piping things together in sequence


### `dplyr::filter()` to keep or exclude rows based on our conditions

1. Exact match (single condition)

Only keep observations where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```

2. "and" statement within filter

Only keep observations if the penguin species is Chinstrap AND the year is 2008
```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)

# ex_2 <- penguins %>% 
#  filter(species == "Chinstrap" & year == 2008)

# ex_2 <- penguins %>% 
#  filter(species == "Chinstrap") %>% 
#  filter(year == 2008)
```

3. Exclude Gentoos

Keep observations if the penguin species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Exercise:

Create a subset from `penguins` that only contains observations for Adelie penguins with a flipper length greater than or equal to 185mm.

```{r}
practice_1 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)
```


4. Using the %in% operator

Use the %in% operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row.

We want to keep observations where the island is Dream or Torgersen.

```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
```


### `select()` to select or exclude COLUMNS

5. Keep a single column by name

Only want to keep species column.
```{r}
ex_5 <- penguins %>% select(species)

# Reminder: rewrite this without the pipe operator
ex_5 <- select(.data = penguins, species)
```

6. Keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year

```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

7. Choose or exclude columns based on position
```{r}
ex_7 <- penguins %>% 
  select(-4)
```

8. A piped sequence

Write a piped sequence where you first: only keep columns for species, body_mass_g, and island; and second: filter to only include observations if body_mass_g is greater than 3600

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>%
  filter(body_mass_g > 3600)
```


### `mutate()` to add or update existing columns, while keeping the ones already there

9. Add a columns that has body mass converted to kg

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000,
         flipper_length_m = flipper_length_mm / 1000)
```


10. Mutate to update an existing column
```{r}
ex_10 <- penguins %>% 
  mutate(island_HOORAY = str_replace(island, pattern = "sen", replacement = "HOORAY!"))
```

11. Update the class of the bill_depth_mm column to character

```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_11$bill_depth_mm)
```

### `case_when()`: your favorite if-else

12. Create a new column that contains "small" if penguin flipper length is less than 185 mm, "medium" otherwise
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```


Task: Starting from penguins, create a new column "bird_bill_bin" that contains:

- If bill length < 38mm: "short bill"
- If bill length >= 38mm & < 43mm: "medium bill"
- If bill length >= 43mm: "long bill"

```{r}
practice_2 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
#   TRUE ~ "long bill" 
  ))
```


### `group_by() %>% summarize()` to find summary values by group, and return them in a nice summary table

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)
```

13. `group_by()` to say there are groups within the species column

```{r}
ex_13 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))
```

## Practice

In a single pipe sequence:

- Only keep penguins that have a body mass > 3500 g
- Only keep columns for species, flipper_length_mm, and sex
- Create a summary table that contains the mean, standard deviation, max and min value of flipper length by species and sex (as the grouping variables)

```{r}
practice_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
#  filter(!is.na(sex)) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flip = mean(flipper_length_mm, na.rm = TRUE),
            sd_flip = sd(flipper_length_mm, na.rm = TRUE),
            min_flip = min(flipper_length_mm, na.rm = TRUE),
            max_flip = max(flipper_length_mm, na.rm = TRUE)) 
#  ungroup() %>% 
#  drop_na(sex)
```












